Cultural Transmission and the Social Contexts of Technological Transitions during the Late Paleolithic of Korea

### 2023-02-24 analysis/paper/paper.qmd
Spelling change for consistency

The application of evolutionary theory to archaeological research has been fruitful for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural]. Archaeologists have used evolutionary theories and methods to get insights into human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current]. In this paper, we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic (40-35 ka). The primary technological innovation of this period was the introduction of stemmed points and blades. These new lithic technologies may represent the first arrival of modern humans in this region, and thus, they may also mark an important event in human dispersal through East Asia [@seong2009emergence]. Previous studies on Korean stemmed points have mostly focused on where these new technologies may have originated, connecting Korea with global patterns of modern human dispersal [@seong2008tanged; @bae2010origin; @bae2017origin]. There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea. Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic, using evolutionary theory to make inferences about cultural transmission from artefact measurements. Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? We also ask: Did modes of cultural transmission vary over time and space? We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two. The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally and independently developed.

@@

Following Bettinger and Eerkens' framework, @garvey2018current used simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is close to the Weber fraction for manually made human artefacts, that is, the threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.

@@

Our correlation analyses found that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias. Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker. Transmission of the artefact shape as a package seems unlikley. The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages, the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage). When all artifacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below. However, further analysis by site and raw material shows more ambiguous patterns, with CV values close to or below the threshold value. Change in CV values over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological Phase 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points, we observed minimal changes in CV values over time (@fig-plot-cv-four-assemblage).

@@

In exploring variation between sites, as a proxy for spatial variation, we found no readily interpretable pattern of variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points. We found that correlation coefficients and CV values are distributed differently in each assemblage. While SYG1_2 and SYG1_3 are from the same site, they belong to different chronological periods, and their correlation decreases over time. It is likely that the correlation or overall shape may be more dependent on individual knappers. We also found that some attributes such as maximum width (MW) and maximum length (ML) have stronger correlation, while the others including stemmed length (SL) and body length (BL) have lower or even negative correlation. Based on these findings, we propose that certain attributes were more carefully transmitted than others. Artefact shape does not appear to have been transmitted as a package, but as a flexible set of attributes that can vary with respect to each other. We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW) (cf. @fig-plot-cv-all-attributes).

>>>>

The application of evolutionary theory to archaeological research has been fruitful for the study of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural]. Archaeologists have used evolutionary theories and methods to get insights into human behavioral ecology, cultural transmission, and artifact phylogenetics in the past [@riede2010isn; @o2017dual; @garvey2018current]. In this paper, we use cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic (40-35 ka). The primary technological innovation of this period was the introduction of stemmed points and blades. These new lithic technologies may represent the first arrival of modern humans in this region, and thus, they may also mark an important event in human dispersal through East Asia [@seong2009emergence]. Previous studies on Korean stemmed points have mostly focused on where these new technologies may have originated, connecting Korea with global patterns of modern human dispersal [@seong2008tanged; @bae2010origin; @bae2017origin]. There are, however, largely unanswered questions about the specific cultural processes and social contexts of this technological change in Korea. Our study explores the social contexts in which new technologies emerged in the Korean Late Paleolithic, using evolutionary theory to make inferences about cultural transmission from artifact measurements. Our main question is: What was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? We also ask: Did modes of cultural transmission vary over time and space? We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two. The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally and independently developed.

@@

Following Bettinger and Eerkens' framework, @garvey2018current used simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying." A CV of 3% is close to the Weber fraction for manually made human artifacts, that is, the threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.

@@

Our correlation analyses found that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias. Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker. Transmission of the artifact shape as a package seems unlikley. The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages, the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage). When all artifacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below. However, further analysis by site and raw material shows more ambiguous patterns, with CV values close to or below the threshold value. Change in CV values over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological Phase 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points, we observed minimal changes in CV values over time (@fig-plot-cv-four-assemblage).

@@

In exploring variation between sites, as a proxy for spatial variation, we found no readily interpretable pattern of variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points. We found that correlation coefficients and CV values are distributed differently in each assemblage. While SYG1_2 and SYG1_3 are from the same site, they belong to different chronological periods, and their correlation decreases over time. It is likely that the correlation or overall shape may be more dependent on individual knappers. We also found that some attributes such as maximum width (MW) and maximum length (ML) have stronger correlation, while the others including stemmed length (SL) and body length (BL) have lower or even negative correlation. Based on these findings, we propose that certain attributes were more carefully transmitted than others. Artifact shape does not appear to have been transmitted as a package, but as a flexible set of attributes that can vary with respect to each other. We observed that CV values of body length (BL), maximum length (ML), mid width (MW), and tang width (TW) are generally lower than tang length (TL), stem length (SL), and stem width (SW) (cf. @fig-plot-cv-all-attributes).
